---
title: "Cooling Center Optimization Utilizing Python SPOPT Library"
output: html_notebook
---

This R-notebook will go through the process of replicating the cooling center optimization in ArcPro using R. Being able to replicate the process in an open source software will enable greater access to the approach be eliminating licensing costs. Additionally, this code will serve as the base for the R-Shiny app which will allow me to present the optimization in an web based interface that decision makers can use directly. 

There seem to be a handful of open source coding libraries I can use to replicate the location modeling I conducted in ArcPro. For now I'm going to attempt two different approaches using a subset of the data I've used in ArcPro. The first approach uses the Maxcovr package in R. The second approach is to use functions in the Pysal libraries in Python. The second approach could be a bit more robust especially with location modeling using a network. However, both tools have been used in the context of a road network. Both approaches use solvers to identify the optimized set of locations to meet demand. For now I'm testing both approaches but in the design of the dashboard I will lean on one approach. 


```{r Setup Python Virtual Environment}
library(tidyverse)
library(reticulate)

## Reticulate to start up your virtual python environment
vrt.env <- "rpython-virtualenv-coolingcenteropt"
use_virtualenv("rpython-virtualenv-coolingcenteropt")
## Install python libraries into virtual environment
python.packages <- c("scipy","spopt","geopandas","pulp","shapely","spaghetti")
virtualenv_install(vrt.env, python.packages)

```

#### Input Data

For the time being, I'm going to read in the facility and demand point data which will be used to calculate the OD matrix which informs the optimization approach. I'm doing some minor editing of the input data to strip away PII and to merge the data into one data frame.
```{r Input_data}

library(dplyr)
library(sf)

candidatefac.file <- "../data/CoolingCenterDashboard-testdata/final_candidate_sites2024PhxUrbanCoresub.shp"
coolingfac.file <- "../data/CoolingCenterDashboard-testdata/Grp18_RespiteCoolingCenters2024_PhxUrbanCoreSub.shp"
homelessdmn.file <- "../data/CoolingCenterDashboard-testdata/Homeless_Outreach_PhxUrbanCoreSub.shp"
residentialdmn.file <- "../data/CoolingCenterDashboard-testdata/Residential_Parcel_Sample_PhxUrbanCoreSub.shp"

candidate.fac <- st_read(candidatefac.file)
cooling.fac <- st_read(coolingfac.file)
homeless.dmn <- st_read(homelessdmn.file)
residential.dmn <- st_read(residentialdmn.file)

## Combine the facility and demand point files into one file before bringing it into Maxcovr Approach
## The combined st object will also include 4 columns: PointID, PointType, Geometry. 

candidate.fac$PointType <- "Candidate Facility"
candidate.fac$PointSource <- "Parcel Dataset"
candidatefac.sub <- candidate.fac[,c("FacilityID","PointType","PointSource","geometry")]
colnames(candidatefac.sub) <- c("PointID","PointType","PointSource","geometry")

cooling.fac$PointType <- "Required Facility"
cooling.fac$PointSource <- "Heat Relief Network"
coolingfac.sub <- cooling.fac[,c("FacilityID","PointType","PointSource","geometry")]
colnames(coolingfac.sub) <- c("PointID","PointType","PointSource","geometry")

homeless.dmn$PointType <- "Demand Point"
homeless.dmn$PointSource <- "Homeless Outreach Survey"
homelessdmn.sub <- homeless.dmn[,c("PointID","PointType","PointSource","geometry")]

residential.dmn$PointType <- "Demand Point"
residential.dmn$PointSource <- "Residential Parcel Dataset"
residentialdmn.sub <- residential.dmn[,c("PointID","PointType","PointSource","geometry")]

points.sf <- rbind(candidatefac.sub,coolingfac.sub,homelessdmn.sub,residentialdmn.sub)
```

#### Maxcovr Approach 

Maxcovr contains several functions that can be useful for location modeling.This includes the maxcovr function itself which, is the function to compute the MCLP. By default maxcovr calculates euclidean distance, which is not helpful for modeling travel along a network. However, the maxcovr function has options to input a distance matrix built on network distances. Additionally, this funciton has been used in this context before. I just need to be careful regarding the use of other functions because many calculate distance from lat lon coordinates. 

With this approach the maxcovr function is used through an optimization solver to identify the optimal solution. In order to run maxcovr we need to calculate the origin-destination (OD) distance matrix for every facility and demand point in our analysis. We are particularly interested in identifying the shortest distances from demand points to facilities. Maxcovr function will use this distance matrix to identify the optimal solution.

I'm currently trying to figure out a way forward for driving the OD matrix in R. I'm currently avoiding Dodger since the distances are weighted based on pedestrian travel. The weighting is designed so that paths suitable for pedestrain travel are preferred but the weighting does result in slightly larger distance measures. Right now I'm trying to see if I can get around the 10,000 point limit for calculated OD matrix via OSRM. If I can set up my own OSRM server or find one other than the default then I should be able to use this approach. 

##### General Steps
1. Read in demand, facility, and road network data
2. Clean data
3. Calculate OD distance matrix between facilities and demand. Ensure network distances are used
4. Run Maxcovr 
5. Review Results

```{r maxcovr-approach}
library(maxcovr)
library(dplyr)
library(sf)


```


### Maxcovr outline from Gemini

```{r}
# 1. Install and load the 'osrm' package
# If you don't have it installed, uncomment the line below and run it:
# install.packages("osrm")
library(osrm)

# 2. Prepare your location data
# Create a data frame with your points.
# Each row represents a location (e.g., a potential facility site or a demand point).
# It must have unique IDs and longitude/latitude coordinates.
# Let's create 5 dummy locations for demonstration.
locations_df <- data.frame(
  id = c("A", "B", "C", "D", "E"),
  lon = c(-123.1814231, -123.25, -123.1, -123.3, -123.15),
  lat = c(49.195113, 49.2, 49.15, 49.25, 49.18)
)

# You can add more locations as needed.
# For Maxcovr, you'll typically have two sets of points:
#   - Candidate facility locations
#   - Demand points (e.g., population centroids)
# For simplicity, this example calculates a matrix between all these 5 points.
# If you have separate origin and destination sets, you'd create two data frames.

# 3. Connect to your local OSRM server
# Set the OSRM server URL. This should match your Docker setup.
# Ensure your OSRM Docker container is running and accessible at this address.
options(osrm.server = "http://127.0.0.1:5000/")

# 4. Calculate the network distance matrix using osrmTable()
# The 'loc' argument takes your data frame of locations.
# The 'measure' argument specifies what you want in the matrix ("distance" for meters, "duration" for seconds).
# The 'osrmTable' function will return a list containing:
#   - $distances: The distance matrix (in meters by default)
#   - $durations: The duration matrix (in seconds by default)
#   - $sources: Information about the source points
#   - $destinations: Information about the destination points

cat("Calculating OSRM distance matrix...\n")
osrm_matrix_result <- osrmTable(loc = locations_df, measure = "distance")

# Extract the distance matrix
distance_matrix <- osrm_matrix_result$distances

# Rename rows and columns for clarity (optional, but good practice)
rownames(distance_matrix) <- locations_df$id
colnames(distance_matrix) <- locations_df$id

cat("\nOSRM Distance Matrix (in meters):\n")
print(distance_matrix)

# You can also get the duration matrix if needed
# duration_matrix <- osrm_matrix_result$durations
# rownames(duration_matrix) <- locations_df$id
# colnames(duration_matrix) <- locations_df$id
# cat("\nOSRM Duration Matrix (in seconds):\n")
# print(duration_matrix)

# 5. Integrate with Maxcovr
# Now that you have 'distance_matrix', you can use it with your Maxcovr function.
# The exact arguments for Maxcovr will depend on how it's defined.
# Below is a conceptual placeholder for how you might use it.

# Assuming 'Maxcovr' is a function that takes:
#   - 'distance_matrix': The network distance matrix
#   - 'demand_points': A vector or data frame of demand points (e.g., population at each location)
#   - 'facility_capacity': The coverage capacity of each facility
#   - 'p': The number of facilities to select

# Dummy data for Maxcovr (replace with your actual data)
# For Maxcovr, you'd typically have a `demand` vector and a `coverage_radius`
# Let's assume demand is 1 for all points and coverage radius is 50000 meters (50 km)
demand_points <- rep(1, nrow(locations_df)) # Assuming equal demand for all locations
names(demand_points) <- locations_df$id
coverage_radius_meters <- 50000 # Example: 50 km coverage radius

# Placeholder for your Maxcovr function call
# If Maxcovr is from a specific package, you'd load that package first.
# For example, if it's from the 'spatstat' or a custom package.

# Example of how Maxcovr might be structured (this is NOT the actual Maxcovr function,
# but illustrates how you'd pass the distance_matrix)
# You would replace this with your actual Maxcovr function call.

# If Maxcovr is a custom function you've defined or from a specific package,
# ensure it's loaded and its parameters match what you're providing.

# For instance, if Maxcovr takes a distance matrix and a coverage threshold:
# selected_facilities <- Maxcovr(
#   distances = distance_matrix,
#   demand = demand_points,
#   coverage_threshold = coverage_radius_meters,
#   num_facilities_to_select = 2 # Example: select 2 facilities
# )

# cat("\nMaxcovr would be called here with the 'distance_matrix'.\n")
# cat("The result would be the selected facility locations based on your criteria.\n")

# Important: Ensure your OSRM Docker container is running before executing this R code.
# You can verify its status by running `docker ps` in your terminal.

```

