---
title: "Cooling Center Optimization Generateing OSRM Matrices"
output: html_notebook
---

This R notebook will contain the code chunks for preparing spatial data and generating the OSRM distance matrices, which will be used in the location modeling later on.  

---
title: OSRM Distance Matrix Generation for Phoenix Cooling Centers
author: Lance Watkins
date: July 22, 2025
output: html_notebook
---

```{r setup, include=FALSE}
# Setup chunk: This runs R code silently when knitting/running notebook
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```


## 1. Project Setup and Data Paths

This section defines the working directory and paths for input/output data. Ensure your Docker OSRM servers are running before executing the OSRM query chunks!

```{r message=TRUE}
# IMPORTANT: Change this path to where your R project is located.
# This directory should contain your .Rmd file and a 'data' subfolder for inputs/outputs.
setwd("C:/Users/raelu/ASU Dropbox/Lance Watkins/ASU/projects/CoolingCenterSpatiaOptimization/CoolingCenterOpt-R-SPOPT")

# Create a 'data' sub-directory if it doesn't exist
if (!dir.exists("data")) {
  dir.create("data")
}

# Define paths for your input CSV files (assuming they are in the 'data' subfolder)
input_data_path <- "data/"
output_data_path <- "data/" # Where to save the generated .rds files

message("Project setup complete. Data paths defined.")
```

## 2. Install and Load Required R Packages

Ensure these packages are installed. If not, uncomment and run install.packages() first.

```{r include=FALSE}
# install.packages(c("sf", "osrm", "dplyr", "lpSolve", "R.utils"))

library(sf)      # For Simple Features (spatial data handling)
library(osrm)    # To interface with the OSRM server
library(dplyr)   # For data manipulation (e.g., mutate, bind_rows)
library(R.utils) # For the chunking function (used within safe_osrmTable)
library(lpSolve) # Will be needed for the MCLP optimization later on
```


## 3. Load and Prepare Spatial Data (Using CSV Files)

This section loads your spatial point data (candidate locations, existing centers, demand points) from CSV files and prepares them for OSRM.

```{r load_prepare_data_csvs, message=TRUE}
message("Loading and preparing spatial data from CSVs...")

# --- IMPORTANT: Ensure your CSV files are in the 'data/' subfolder ---
# And check that they have the required columns: 'id', 'latitude', 'longitude'.
# For demand_points.csv, also ensure a 'population' or 'weight' column.

# Load Candidate Locations
candidate_locs <- read.csv(paste0(input_data_path, "candidate_locations.csv")) %>%
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>% # crs 4326 is WGS84 (lat/lon)
  # Ensure unique 'id' column is of character type. Adjust 'id' if your CSV column is named differently.
  dplyr::mutate(id = as.character(id))

# Load Existing Cooling Center Locations
existing_locs <- read.csv(paste0(input_data_path, "existing_cooling_centers.csv")) %>%
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  # Ensure unique 'id' column is of character type. Adjust 'id' if your CSV column is named differently.
  dplyr::mutate(id = as.character(id)) %>%
  # Add an Address column so the columns align with candidate locations dataframe as well. For now populate with NA
  dplyr::mutate(address = NA) %>%
  dplyr::select(c("id","address","type","geometry"))

# Load Demand Points
demand_pts <- read.csv(paste0(input_data_path, "demand_points.csv")) %>%
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  # Ensure unique 'id' column is of character type. Adjust 'id' if your CSV column is named differently.
  dplyr::mutate(id = as.character(id))



# Combine all potential facility locations (candidate + existing)
# This combined set will be the 'sources' for your OSRM queries.

all_facilities <- dplyr::bind_rows(candidate_locs, existing_locs) %>%
  # Add a simple numeric_id for consistent internal matrix indexing,
  # although we will primarily use the 'id' column for clarity in results.
  dplyr::mutate(numeric_id = dplyr::row_number())

# Optional: Basic check for unique IDs (good practice)
if (length(unique(all_facilities$id)) != nrow(all_facilities)) {
  warning("Duplicate 'id' values found in combined all_facilities. This might cause issues with matrix naming.")
}
if (length(unique(demand_pts$id)) != nrow(demand_pts)) {
  warning("Duplicate 'id' values found in demand_pts. This might cause issues with matrix naming.")
}

message("Spatial data loaded and prepared. Sample of all_facilities:")
print(head(all_facilities))
message("Sample of demand_pts:")
print(head(demand_pts))
```


## 4. Define Helper Function for OSRM Queries

## 5. Calculate Network Distance Matrices using OSRM


