---
title: "Cooling Center Optimization Generateing OSRM Matrices"
output: html_notebook
---

This R notebook will contain the code chunks for preparing spatial data and generating the OSRM distance matrices, which will be used in the location modeling later on.  

---
title: "OSRM Distance Matrix Generation for Phoenix Cooling Centers"
author: "Your Name"
date: "July 22, 2025"
output: html_notebook
---

```{r setup, include=FALSE}
# Setup chunk: This runs R code silently when knitting/running notebook
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```


1. Project Setup and Data Paths

This section defines the working directory and paths for input/output data. Ensure your Docker OSRM servers are running before executing the OSRM query chunks!

```{r}
# IMPORTANT: Change this path to where your R project is located.
# This directory should contain your .Rmd file and a 'data' subfolder for inputs/outputs.
setwd("C:/Users/raelu/ASU Dropbox/Lance Watkins/ASU/projects/CoolingCenterSpatiaOptimization/CoolingCenterOpt-R-SPOPT")

# Create a 'data' sub-directory if it doesn't exist
if (!dir.exists("data")) {
  dir.create("data")
}

# Define paths for your input CSV files (assuming they are in the 'data' subfolder)
input_data_path <- "data/"
output_data_path <- "data/" # Where to save the generated .rds files

message("Project setup complete. Data paths defined.")
```

2. Install and Load Required R Packages

Ensure these packages are installed. If not, uncomment and run install.packages() first.

```{r}
# install.packages(c("sf", "osrm", "dplyr", "lpSolve", "R.utils"))

library(sf)      # For Simple Features (spatial data handling)
library(osrm)    # To interface with the OSRM server
library(dplyr)   # For data manipulation (e.g., mutate, bind_rows)
library(R.utils) # For the chunking function (used within safe_osrmTable)
library(lpSolve) # Will be needed for the MCLP optimization later on
```









